Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-09-27 paper.tex
Deploying to gh-pages from @ bautheac/financialisation-co-movement@c328f0b9c5a755b03d340f3f58900773e4ce888e 🚀

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the UK metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME). This sample cross-section of assets is further divided into
groups for the purpose of the analysis by country (US, GB), sector
(agriculturals, energy, metals) and by sub-sector (agriculturals:
grains, livestock, softs; energy: gas, petroleum; metals: base,
precious).

@@

Table~\ref{tbl-stats-no-regimes} displays descriptive statistics for the

@@

\textbf{CHP} is low and vice et versa. In Table~\ref{tbl-stats-regimes}
the mean returns are shown over high (\textbf{contango}) and low
(\textbf{backwardation}) \textbf{CHP} regimes for each time period.\\

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a
unified asset class with interconnectedness emerging as a crucial aspect
that warrants thorough examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicate higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-regimes}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years} to uncover any distinct breakpoints.
For the full sample of assets, the average pairwise correlations
increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US
metals average pairwise correlations, more than doubling from 0.22 to
0.49, suggesting 2004 as the onset of financialization as noted in
several studies \citep{baker_financialization_2021, tang_index_2012} and
our contribution is to observe that the US metals sub-sector seems to
have played a major role in this.

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-no-regimes}) with eighteen US commodities along
with all six GB traded metals showing lower mean returns. Returns for
equally weighted portfolios of both sets of assets fall down to 9.4\%
and 5.3\% from 15.9\% and 21.8\% respectively. Returns are mostly higher
during phases of \textbf{backwardation} relative to \textbf{contango}
(Table~\ref{tbl-stats-regimes}) with fifteen US assets and five of the
GB traded metals showing higher mean returns in \textbf{backwardation}
and the pattern extending to equally weighted portfolios of the two sets
of assets. Volatility is moderately higher for both sets with fifteen US
commodities and three GB metals showing higher figures and interestingly
eminently higher over phases of \textbf{contango}; out of the US complex
only Sugar-\#11 (IFUS) and none of the GB traded metals show higher
volatility figures over \textbf{backwardation} phases. The symmetry
between the dynamics of CHP and \textbf{CHP}'s explanatory power on
individual commodity returns breaks in the crisis period
(Table~\ref{tbl-time-reg-US-CHP}) with both now seemingly converging;
the US cross-sectional average \(R^{2}\) at the individual level
bouncing back up (22.5\% vs 20.8\%) while that at the aggregate level
rises further (14.7\% vs 7.1\%).

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-no-regimes}) with twenty-two US and four UK traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all UK metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and UK assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all UK metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and UK assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period (Table~\ref{tbl-stats-regimes}) with mean returns
higher during phases of \textbf{backwardation} relative to
\textbf{contango}, more so than over the crisis period with seventeen US
commodities showing higher mean returns over \textbf{backwardation}
phases. The difference is significant at the 5\% level for Lumber (XCME)
and at the 10\% level for Palladium (XNYM) while no difference is
significant for the assets showing higher figures over phases of
\textbf{contango}. The pattern is more pronounced for the UK traded
subset with all six assets showing higher mean returns in
\textbf{backwardation} with the difference significant at the 1\% level
for Nickel-primary (XLME), 5\% level for Aluminium (XLME) and 10\% level
for Zinc (XLME). It also extends to equally weighted portfolios of US
and UK traded assets that both show higher mean returns during
\textbf{backwardation} with the difference significant at the 5\% level
for the UK portfolio and the figure close to the 10\% level significance
threshold for the portfolio of US assets. Volatilities are overall
higher in \textbf{contango} although less so than over the previous
period with seventeen US commodities showing higher figures over
\textbf{contango} against twenty-three in the crisis period. The pattern
extends to three of the six UK traded metals with the difference
significant at the 1\% level for Copper (XLME) and Tin (XLME) and at the
5\% level for Zinc (XLME) while differences are not significant for
Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME) that exhibit
higher volatility over \textbf{backwardation} phases. Equally weighted
portfolios of US and UK traded assets on the other hand both show higher
volatility in \textbf{contango} with the difference significant at the
1\% level for both. Over the financialisation period in contrast
volatilities were predominantly higher in \textbf{backwardation} for US
traded assets with fourteen commodities showing higher volatility
figures in these phases; the post-crisis pattern of results for
volatilities thus likewise shows signs of reversion to
pre-financialisation standards. After showing signs of convergence in
the crisis period the dynamics of the explanatory power of CHP and
\textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}), seemingly
reverting to the Keynesian paradigm observed pre-financialisation,
further suggesting a return to legacy commodity specific dynamics. The
recovery initiated in the crisis period for that of CHP is confirmed
with a stronger increase in the post-crisis period, 27.2\% vs.~22.5\%,
strongest for the metals sector, 37.3\% vs.~23.7\%, while the energy
sector shows the opposite pattern: 11\% vs.~16\%. In contrast, after
increasing stronger in the crisis, that for \textbf{CHP} falls overall
in the post-crisis period, 7.5\% vs 14.7\%, with the decrease strongest
for the energy sector in a relative sense: 2.5\% vs.~14.8\%.

@@

\caption{\label{tbl-stats-combined}This table shows mean returns and
volatility (sd) for the twenty four individual US commodities and the
six GB metals considered in the study as well as for two equally
weighted portfolios formed from the US commodities and the GB metals
respectively across the four periods of interest (past: 1997-2003;
financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018).
for each period the figures are shown independently for the entire
period as well as for phases of aggregate backwardation (aggregate CHP
\(\leq\) period median) and aggregate contango (aggregate CHP
\textgreater{} period median). Mean values significant at the 1\%, 5\%
and 10\% level are marked with ***, ** and * respectively. Aggregate CHP
construction and corresponding regime definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results}.}

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{23\%}} & \textcolor[HTML]{4285f4}{\textbf{29.54\%}} & \textcolor[HTML]{4285f4}{\textbf{36.15\%}} & \textcolor[HTML]{4285f4}{\textbf{21.8\%}}\\

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{23.07\%}} & \textcolor[HTML]{4285f4}{\textbf{29.14\%}} & \textcolor[HTML]{4285f4}{\textbf{31.59\%}} & \textcolor[HTML]{4285f4}{\textbf{21\%}}\\

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{23\%}} & \textcolor[HTML]{4285f4}{\textbf{30.01\%}} & \textcolor[HTML]{4285f4}{\textbf{40.14\%}} & \textcolor[HTML]{4285f4}{\textbf{22.68\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{32.22\%}} & \textcolor[HTML]{4285f4}{\textbf{31.04\%}} & \textcolor[HTML]{4285f4}{\textbf{34.67\%}} & \textcolor[HTML]{4285f4}{\textbf{29.5\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{29.02\%}} & \textcolor[HTML]{4285f4}{\textbf{33.82\%}} & \textcolor[HTML]{4285f4}{\textbf{31.43\%}} & \textcolor[HTML]{4285f4}{\textbf{31.55\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{35.22\%}} & \textcolor[HTML]{4285f4}{\textbf{28.1\%}} & \textcolor[HTML]{4285f4}{\textbf{37.7\%}} & \textcolor[HTML]{4285f4}{\textbf{27.5\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{26.34\%}} & \textcolor[HTML]{4285f4}{\textbf{32.55\%}} & \textcolor[HTML]{4285f4}{\textbf{31.78\%}} & \textcolor[HTML]{4285f4}{\textbf{25.11\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{27.77\%}} & \textcolor[HTML]{4285f4}{\textbf{33.06\%}} & \textcolor[HTML]{4285f4}{\textbf{25.7\%}} & \textcolor[HTML]{4285f4}{\textbf{25.1\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{24.94\%}} & \textcolor[HTML]{4285f4}{\textbf{32.13\%}} & \textcolor[HTML]{4285f4}{\textbf{36.85\%}} & \textcolor[HTML]{4285f4}{\textbf{25.32\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.43\%}} & \textcolor[HTML]{4285f4}{\textbf{28.35\%}} & \textcolor[HTML]{4285f4}{\textbf{25.75\%}} & \textcolor[HTML]{4285f4}{\textbf{18.95\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.35\%}} & \textcolor[HTML]{4285f4}{\textbf{28.94\%}} & \textcolor[HTML]{4285f4}{\textbf{22.33\%}} & \textcolor[HTML]{4285f4}{\textbf{18.19\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.55\%}} & \textcolor[HTML]{4285f4}{\textbf{27.91\%}} & \textcolor[HTML]{4285f4}{\textbf{28.65\%}} & \textcolor[HTML]{4285f4}{\textbf{19.79\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.47\%}} & \textcolor[HTML]{4285f4}{\textbf{30.04\%}} & \textcolor[HTML]{4285f4}{\textbf{28\%}} & \textcolor[HTML]{4285f4}{\textbf{21.22\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.93\%}} & \textcolor[HTML]{4285f4}{\textbf{34.58\%}} & \textcolor[HTML]{4285f4}{\textbf{24.39\%}} & \textcolor[HTML]{4285f4}{\textbf{22.04\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{21.09\%}} & \textcolor[HTML]{4285f4}{\textbf{24.74\%}} & \textcolor[HTML]{4285f4}{\textbf{31.11\%}} & \textcolor[HTML]{4285f4}{\textbf{20.56\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{24.93\%}} & \textcolor[HTML]{4285f4}{\textbf{33.03\%}} & \textcolor[HTML]{4285f4}{\textbf{37.62\%}} & \textcolor[HTML]{4285f4}{\textbf{26.58\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{25.79\%}} & \textcolor[HTML]{4285f4}{\textbf{36\%}} & \textcolor[HTML]{4285f4}{\textbf{37.53\%}} & \textcolor[HTML]{4285f4}{\textbf{26.57\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{contango} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{24.04\%}} & \textcolor[HTML]{4285f4}{\textbf{29.84\%}} & \textcolor[HTML]{4285f4}{\textbf{37.73\%}} & \textcolor[HTML]{4285f4}{\textbf{26.72\%}}\\

@@

\textbf{Cattle-feeder (XCME)} & \textbf{whole period} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{13.23\%}} & \textcolor[HTML]{4285f4}{\textbf{15.16\%}} & \textcolor[HTML]{4285f4}{\textbf{15.24\%}} & \textcolor[HTML]{4285f4}{\textbf{18.69\%}}\\

@@

\textbf{Cattle-feeder (XCME)} & \textbf{backwardation} & \textbf{sd} & \textcolor[HTML]{4285f4}{\textbf{12.79\%}} & \textcolor[HTML]{4285f4}{\textbf{14.93\%}} & \textcolor[HTML]{4285f4}{\textbf{13.39\%}} & \textcolor[HTML]{4285f4}{\textbf{20.14\%}}\\

@@

>>>>

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the UK metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME). This sample cross-section of assets is further divided into
groups for the purpose of the analysis by country (US, GB), sector
(agriculturals, energy, metals) and by sub-sector (agriculturals:
grains, livestock, softs; energy: gas, petroleum; metals: base,
precious)\footnote{See Section~\ref{sec-taxonomy} for a detailed
  taxonomy of the commodity assets considered in the study.}.

@@

Table~\ref{tbl-stats-combined} displays descriptive statistics for the

@@

\textbf{CHP} is low and vice et versa. Table~\ref{tbl-stats-combined}
further shows the mean returns and volatilities for high
(\textbf{contango}) and low (\textbf{backwardation}) \textbf{CHP}
regimes for each time period.\\

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a
unified asset class with interconnectedness emerging as a crucial aspect
that warrants thorough examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicate higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years} to uncover any distinct breakpoints.
For the full sample of assets, the average pairwise correlations
increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US
metals average pairwise correlations, more than doubling from 0.22 to
0.49, suggesting 2004 as the onset of financialization as noted in
several studies \citep{baker_financialization_2021, tang_index_2012} and
our contribution is to observe that the US metals sub-sector seems to
have played a major role in this.

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-combined}) with eighteen US commodities along with
all six GB traded metals showing lower mean returns. Returns for equally
weighted portfolios of both sets of assets fall down to 9.4\% and 5.3\%
from 15.9\% and 21.8\% respectively. Returns are mostly higher during
phases of \textbf{backwardation} relative to \textbf{contango} with
fifteen US assets and five of the GB traded metals showing higher mean
returns in \textbf{backwardation} and the pattern extending to equally
weighted portfolios of the two sets of assets. Volatility is moderately
higher for both sets with fifteen US commodities and three GB metals
showing higher figures and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks in the crisis period (Table~\ref{tbl-time-reg-US-CHP}) with both
now seemingly converging; the US cross-sectional average \(R^{2}\) at
the individual level bouncing back up (22.5\% vs 20.8\%) while that at
the aggregate level rises further (14.7\% vs 7.1\%).

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-combined}) with twenty-two US and four UK traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all UK metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and UK assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all UK metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and UK assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period with mean returns higher during phases of
\textbf{backwardation} relative to \textbf{contango}, more so than over
the crisis period with seventeen US commodities showing higher mean
returns over \textbf{backwardation} phases. The difference is
significant at the 5\% level for Lumber (XCME) and at the 10\% level for
Palladium (XNYM) while no difference is significant for the assets
showing higher figures over phases of \textbf{contango}. The pattern is
more pronounced for the UK traded subset with all six assets showing
higher mean returns in \textbf{backwardation} with the difference
significant at the 1\% level for Nickel-primary (XLME), 5\% level for
Aluminium (XLME) and 10\% level for Zinc (XLME). It also extends to
equally weighted portfolios of US and UK traded assets that both show
higher mean returns during \textbf{backwardation} with the difference
significant at the 5\% level for the UK portfolio and the figure close
to the 10\% level significance threshold for the portfolio of US assets.
Volatilities are overall higher in \textbf{contango} although less so
than over the previous period with seventeen US commodities showing
higher figures over \textbf{contango} against twenty-three in the crisis
period. The pattern extends to three of the six UK traded metals with
the difference significant at the 1\% level for Copper (XLME) and Tin
(XLME) and at the 5\% level for Zinc (XLME) while differences are not
significant for Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME)
that exhibit higher volatility over \textbf{backwardation} phases.
Equally weighted portfolios of US and UK traded assets on the other hand
both show higher volatility in \textbf{contango} with the difference
significant at the 1\% level for both. Over the financialisation period
in contrast volatilities were predominantly higher in
\textbf{backwardation} for US traded assets with fourteen commodities
showing higher volatility figures in these phases; the post-crisis
pattern of results for volatilities thus likewise shows signs of
reversion to pre-financialisation standards. After showing signs of
convergence in the crisis period the dynamics of the explanatory power
of CHP and \textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}),
seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a return to legacy commodity
specific dynamics. The recovery initiated in the crisis period for that
of CHP is confirmed with a stronger increase in the post-crisis period,
27.2\% vs.~22.5\%, strongest for the metals sector, 37.3\% vs.~23.7\%,
while the energy sector shows the opposite pattern: 11\% vs.~16\%. In
contrast, after increasing stronger in the crisis, that for \textbf{CHP}
falls overall in the post-crisis period, 7.5\% vs 14.7\%, with the
decrease strongest for the energy sector in a relative sense: 2.5\%
vs.~14.8\%.

@@

\caption{\label{tbl-stats-combined}This table shows mean returns and
volatility for the twenty four individual US commodities and the six GB
metals considered in the study as well as for two equally weighted
portfolios formed from the US commodities and the GB metals respectively
across the four periods of interest (past: 1997-2003; financialisation:
2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). for each period
the figures are shown independently for the entire period as well as for
phases of aggregate backwardation (aggregate CHP \(\leq\) period median)
and aggregate contango (aggregate CHP \textgreater{} period median).
Mean values significant at the 1\%, 5\% and 10\% level are marked with
***, ** and * respectively. Aggregate CHP construction and corresponding
regime definitions are discussed in Section~\ref{sec-methods} while the
results are discussed in Section~\ref{sec-results}.}

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{23\%}} & \textcolor[HTML]{4285f4}{\textbf{29.54\%}} & \textcolor[HTML]{4285f4}{\textbf{36.15\%}} & \textcolor[HTML]{4285f4}{\textbf{21.8\%}}\\

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{23.07\%}} & \textcolor[HTML]{4285f4}{\textbf{29.14\%}} & \textcolor[HTML]{4285f4}{\textbf{31.59\%}} & \textcolor[HTML]{4285f4}{\textbf{21\%}}\\

@@

\textbf{Corn-\#2 yellow (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{23\%}} & \textcolor[HTML]{4285f4}{\textbf{30.01\%}} & \textcolor[HTML]{4285f4}{\textbf{40.14\%}} & \textcolor[HTML]{4285f4}{\textbf{22.68\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{32.22\%}} & \textcolor[HTML]{4285f4}{\textbf{31.04\%}} & \textcolor[HTML]{4285f4}{\textbf{34.67\%}} & \textcolor[HTML]{4285f4}{\textbf{29.5\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{29.02\%}} & \textcolor[HTML]{4285f4}{\textbf{33.82\%}} & \textcolor[HTML]{4285f4}{\textbf{31.43\%}} & \textcolor[HTML]{4285f4}{\textbf{31.55\%}}\\

@@

\textbf{Oats (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{35.22\%}} & \textcolor[HTML]{4285f4}{\textbf{28.1\%}} & \textcolor[HTML]{4285f4}{\textbf{37.7\%}} & \textcolor[HTML]{4285f4}{\textbf{27.5\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{26.34\%}} & \textcolor[HTML]{4285f4}{\textbf{32.55\%}} & \textcolor[HTML]{4285f4}{\textbf{31.78\%}} & \textcolor[HTML]{4285f4}{\textbf{25.11\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{27.77\%}} & \textcolor[HTML]{4285f4}{\textbf{33.06\%}} & \textcolor[HTML]{4285f4}{\textbf{25.7\%}} & \textcolor[HTML]{4285f4}{\textbf{25.1\%}}\\

@@

\textbf{Soybean meal (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{24.94\%}} & \textcolor[HTML]{4285f4}{\textbf{32.13\%}} & \textcolor[HTML]{4285f4}{\textbf{36.85\%}} & \textcolor[HTML]{4285f4}{\textbf{25.32\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.43\%}} & \textcolor[HTML]{4285f4}{\textbf{28.35\%}} & \textcolor[HTML]{4285f4}{\textbf{25.75\%}} & \textcolor[HTML]{4285f4}{\textbf{18.95\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.35\%}} & \textcolor[HTML]{4285f4}{\textbf{28.94\%}} & \textcolor[HTML]{4285f4}{\textbf{22.33\%}} & \textcolor[HTML]{4285f4}{\textbf{18.19\%}}\\

@@

\textbf{Soybean oil (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.55\%}} & \textcolor[HTML]{4285f4}{\textbf{27.91\%}} & \textcolor[HTML]{4285f4}{\textbf{28.65\%}} & \textcolor[HTML]{4285f4}{\textbf{19.79\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.47\%}} & \textcolor[HTML]{4285f4}{\textbf{30.04\%}} & \textcolor[HTML]{4285f4}{\textbf{28\%}} & \textcolor[HTML]{4285f4}{\textbf{21.22\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.93\%}} & \textcolor[HTML]{4285f4}{\textbf{34.58\%}} & \textcolor[HTML]{4285f4}{\textbf{24.39\%}} & \textcolor[HTML]{4285f4}{\textbf{22.04\%}}\\

@@

\textbf{Soybeans (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{21.09\%}} & \textcolor[HTML]{4285f4}{\textbf{24.74\%}} & \textcolor[HTML]{4285f4}{\textbf{31.11\%}} & \textcolor[HTML]{4285f4}{\textbf{20.56\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{24.93\%}} & \textcolor[HTML]{4285f4}{\textbf{33.03\%}} & \textcolor[HTML]{4285f4}{\textbf{37.62\%}} & \textcolor[HTML]{4285f4}{\textbf{26.58\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{25.79\%}} & \textcolor[HTML]{4285f4}{\textbf{36\%}} & \textcolor[HTML]{4285f4}{\textbf{37.53\%}} & \textcolor[HTML]{4285f4}{\textbf{26.57\%}}\\

@@

\textbf{Wheat-SRW (XCBT)} & \textbf{contango} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{24.04\%}} & \textcolor[HTML]{4285f4}{\textbf{29.84\%}} & \textcolor[HTML]{4285f4}{\textbf{37.73\%}} & \textcolor[HTML]{4285f4}{\textbf{26.72\%}}\\

@@

\textbf{Cattle-feeder (XCME)} & \textbf{whole period} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{13.23\%}} & \textcolor[HTML]{4285f4}{\textbf{15.16\%}} & \textcolor[HTML]{4285f4}{\textbf{15.24\%}} & \textcolor[HTML]{4285f4}{\textbf{18.69\%}}\\

@@

\textbf{Cattle-feeder (XCME)} & \textbf{backwardation} & \textbf{volatility} & \textcolor[HTML]{4285f4}{\textbf{12.79\%}} & \textcolor[HTML]{4285f4}{\textbf{14.93\%}} & \textcolor[HTML]{4285f4}{\textbf{13.39\%}} & \textcolor[HTML]{4285f4}{\textbf{20.14\%}}\\

@@

\section{Appendix}\label{appendix}

\subsection{Commodity assets taxonomy}\label{sec-taxonomy}
